Notes of Basic Topolgy

### 2016-12-15 Basic Topology Notes/sections/ch6-Triangulation.tex
[major] Basic Topology: Get the generators and relations in the Edge group.

Let $X$ be a path-connected triangulable space, take a specific
triangulation $h:|K|\to X$, and replace $X$ by $|K|$ (we are at
liberty to do this since the funda- mental group is a topological
invariant). Now the advantage of a polyhedron $|K|$ is that the
elements of its fundamental group can be represented by loops which
are made up of edges of $K$. Using such 'edge loops' we shall
construct a group, called the edge group of the complex $|K|$, which
can be computed and wh ich is isomorphie to the fundamental group of
$|K|$.
} \end{myquote}

@@

\begin{thm}
    $E(K,v)$ is isomorphie to $\pi_1(|K|,v)$
\end{thm}
\begin{proof}
    The proof can be found on page 133 of \cite{book}. The essential
    idea is that an edge loop is just a loop. And a loop is just a
    continuous map $f:I\to |K|$, we can easily use the simplicial
    approximation $s:I^m\to |K|$ of $f$ (with appropriately $m$), to
    find a corresponding edge loop.

    The uniqueness of the loop corresponding to an edge loop is found
    again by treating $L:=I^2$ as a simplicial complex and apprximate
    again the homotopy $F:L\to |K|$. The resulted $S$ will again gives
    us a equivalence relation to the constant edge loop. (Note, we
    have to use the fact the simplicial maps preserves the equivalence
    relation between edge pathes). The following pictorially
    produces the proof:
    \begin{figure}[H]
        \centering
        \includegraphics[width=1\linewidth]{pics/ch6-notes-2/L.png}
    \end{figure}
\end{proof}

>>>>

Let $X$ be a path-connected triangulable space, take a specific
triangulation $h:|K|\to X$, and replace $X$ by $|K|$ (we are at
liberty to do this since the funda- mental group is a topological
invariant). Now the advantage of a polyhedron $|K|$ is that the
elements of its fundamental group can be represented by loops which
are made up of edges of $K$. Using such 'edge loops' we shall
construct a group, called the edge group of the complex $|K|$, which
can be computed and wh ich is isomorphic to the fundamental group of
$|K|$.
} \end{myquote}

@@

\begin{thm}
    $E(K,v)$ is isomorphic to $\pi_1(|K|,v)$
\end{thm}
\begin{proof}
    The proof can be found on page 133 of \cite{book}. The essential
    idea is that an edge loop is just a loop. And a loop is just a
    continuous map $f:I\to |K|$, we can easily use the simplicial
    approximation $s:I^m\to |K|$ of $f$ (with appropriately $m$), to
    find a corresponding edge loop.

    The uniqueness of the loop corresponding to an edge loop is found
    again by treating $L:=I^2$ as a simplicial complex and apprximate
    again the homotopy $F:L\to |K|$. The resulted $S$ will again gives
    us a equivalence relation to the constant edge loop. (Note, we
    have to use the fact the simplicial maps preserves the equivalence
    relation between edge pathes). The following pictorially
    produces the proof:
    \begin{figure}[H]
        \centering
        \includegraphics[width=0.9\linewidth]{pics/ch6-notes-2/L.png}
    \end{figure}
\end{proof}

\begin{remark}
    This theorem already tells us that the Edge group $E(K,v)$ is
    indepedent on the choice of base point $v$, since our polyhedron
    $|K|$ is always assumed to be path-connected.
\end{remark}

Based on this edge group, we will try to calculate more explicitly the
structure of this group. Here we provide a way to get the generators
and relations directly from the edge group.

\begin{key}
    The intuition is that the structure of this group is
    determined only by those $1$-simplexes, $2$-simplexes, and the
    equivalence relations between edge pathes. All higher dimensional
    simplexes are not relevent to this group structure.
\end{key}

We let \nomen{$L$} the subcomplex of a simplicial complex $K$. This
subcomplex $L$ is required to contain all the vertices of $K$ and such
that its polyhedron $|L|$ is path-connected and simply-connected. Can
we find such a subcomplex? The intuitive answer is yes due to the
path-connectedness nature of $|K|$. The rigorous answer needs the help
of a concept \nomen{tree}. A tree $T$ is a one-dimensional subcomplex of
$K$ whose polyhedron $|T|$ is both path-connected and
simply-connected. The following lemma affirms that we can find our $L$
to be exactly the maximal tree in $K$:
\begin{lemma}
    A maximal tree contains all the vertices of $K$.
\end{lemma}
\begin{proof}
    We first observe that a tree must exists, unless the dimension of
    $K$ is $0$ (in which case, there is entirely no need for the
    following discussions, and the lemma is still true because there is
    not even a maximal tree!). Let $T$ be a maximal tree in $K$, which
    means that if $T'$ is a tree and contains $T$, than $T'=T$. 
    The next step is simply and is expatiated on page 134 of
    \cite{book}. The idea is to grow a tree by adding more vertices,
    which is possible because $|K|$ is path-connected and a path can
    be easily turned into an edge path. Note that add a
    simply-connected space in this way will keep the space
    simply-connected, just do not add a vertices twice too creat a
    loop.

    Obviously one will get a tree which is maximal and contains all
    the vertices.  Such a process also affirms that the maximal tree
    will exist (but may not be unique).
\end{proof}

\begin{ex}
    Here is the maximal tree of the triangulation of two-torus:
    \begin{figure}[H]
        \centering
        \includegraphics[width=0.9\linewidth]{pics/ch6-notes-3/ex1-maximal-tree.png}
    \end{figure}
\end{ex}

\begin{ex}
    The maximal tree is usually not unique!
    \begin{figure}[H]
        \centering
        \includegraphics[width=0.9\linewidth]{pics/ch6-notes-3/ex2-maxi-tree-not-unique.png}
    \end{figure}
\end{ex}

Now since $|L|$ is simply-connected, it will not contribute to the
edge group. What contributes to the edge group is those edges that are
not in this $|L|$. This motivates us to define a new group:

\begin{defi}[$G(K,L)$]
\nomenclature{$G(K,L)$}{\nomrefpage.}
    Let $L$ be the maximal tree in $K$ (assume $|K|$ is
    path-connected). Then $G(K,L)$ is defined by generators $g_{ij}$
    for each \textbf{ordered} pair of vertices $v_i,v_j$ such that
    $\{v_i,v_j\}$ spans a simplex ($0$-simplex or $1$-simplex) in $K$.
    Those generators are subject to the following conditions:
    \begin{enumerate}
        \item $g_{ij}=1$, the identity element, if and only if
            $\{v_i,v_j\}$ spans a simplex in $L$.
        \item $g_{ij}g_{jk}=g_{ik}$ if and only if $\{v_i,v_j,v_k\}$
            spans a simplex ($0$ or $1$ or $2$ simplex) in $K$.
    \end{enumerate}
\end{defi}

\begin{ex}
    \label{ex:torus-GLKgroup}
    Again, for torus:
    \begin{figure}[H]
        \centering
        \includegraphics[width=0.9\linewidth]{pics/ch6-notes-3/ex3-L-in-triangle.png}
    \end{figure}
\end{ex}
\begin{ex}
    \label{ex:ex4-L-in-two-torus}
    And for another:
    \begin{figure}[H]
        \centering
        \includegraphics[width=0.9\linewidth]{pics/ch6-notes-3/ex4-L-in-two-torus.png}
    \end{figure}
    Note that this does not gives a triangulation of two-torus! 
    The reason will be explained later in
    example~\ref{ex:pi1-two-torus}.
\end{ex}
More generally, we have
\begin{thm}
    $G(K,L)\cong E(K,v)$, as a group isomorphism.
\end{thm}
\begin{proof}
    The proof in page 135 to 136 of \cite{book} constructs the
    following homomorphisms:

    $$\begin{tikzcd}[]
        G(K,L)\ar[r,"\phi",shift left] & 
            E(K,v)\ar[l,"\theta", shift left]
    \end{tikzcd}$$

    It is pretty simple to get these two maps and confirms that their
    composites are identity functions on each space.
\end{proof}

\begin{remark}
    We have now the following chain of homomorphisms:
    $$ \pi_1(|K|,v)\cong E(K,v) \cong G(K,L) $$
\end{remark}
\begin{coro}
    For any path-connected, triangulable space, its fundamental group
    is finitely presented, i.e., the fundamental group has only a
    finite number of generators and a finite number of relations.
\end{coro}

For example, we have demonstrated in example~\ref{ex:torus-GLKgroup}
that its $G(K,L)\cong \Z_2$, which is just its fundamental group.

In addition, armed with this tool, we can answer an additional
question. We want to know what is the fundamental group of two torus
connected together:

\begin{figure}[H]
    \centering
    \includegraphics[width=0.9\linewidth]{pics/ch6-notes-3/pic-two-torus.png}
\end{figure}

This can be done using the following theorem:

\begin{thm}[Van Kampen's theorem]
    Let $J$, $K$ be two simplicial complexes in some Euclidean space
    which intersect in a common subcomplex, and suppose all $|J|$,
    $|K|$, $|J\cap K|$ are path-connected. Let $j,k$ be the inclusion
    maps:
    $$ \begin{tikzcd}[]
        \lvert J\cap K\rvert \ar[r,"j",hook] & \lvert J\rvert
    \end{tikzcd}$$
    $$ \begin{tikzcd}[]
        \lvert J\cap K\rvert \ar[r,"k",hook] & \lvert K\rvert
    \end{tikzcd}$$
    Take a vertice $v\in J\cap K$ as the base point. The fundamental
    group $\pi_1(|J\cup K|, v)$ is obtained from the free product
    $\pi_1(|J|,v)*\pi_1(|K|,v)$ by adding the relations
    $j_*(z)=k_*(z)$ for all $z\in\pi_1(|J\cap K|,v)$.
\end{thm}
\begin{proof}
    The proof can be found on page 137 to 138 of \cite{book}. Please
    do read the paragraph just before this theorem to get a picture of
    the proof.

    Note also that a $2$-simplex cannot possiblly cross the
    intersection part $J\cap K$. Please check why I mention this when
    reading the proof on the book \cite{book}.
\end{proof}

\begin{ex}
    \label{ex:pi1-two-torus}
    Here's how to find the fundamental group for two-torus connected:
\begin{figure}[H]
    \centering
    \includegraphics[width=0.9\linewidth]{pics/ch6-notes-3/ex-pi1-two-torus.png}
\end{figure}
    The result confirms that the triangulation of a two-torus is as
    simple as in example~\ref{ex:ex4-L-in-two-torus}.
\end{ex}

\begin{ex}
    Next we consider the projective space $\R P^2$, the space is
    homeomorphic to half the sphere $S^2$, with the antipodal points
    on the boundary identified:
    \begin{figure}[H]
        \centering
        \includegraphics[width=0.4\linewidth]{pics/ch6-notes-3/rp2.png}
    \end{figure}
    Then it can be regarded as a 2-disc with it boundary attached to
    the boundary of a M\"obius strip. This construction will tell us
    that the fundamental group of $\R P^2$ is $\Z_2$.
    \begin{figure}[H]
        \centering
        \includegraphics[width=0.8\linewidth]{pics/ch6-notes-3/rp2-pi1.png}
        \caption{Page 139, Fig 6.16 of \cite{book}}
    \end{figure}
\end{ex}
